Title Goes Here

### 2021-01-08 analysis/paper/paper.Rmd
Agrega SAM datos y más (🥐️ para charles)

lab_lev <- AddSuffix(" hPa")
lab_cplx <- c(R = "Real", 
              I = "Imaginary")
combine_lists <- function(...) {
  # browser(expr = length(x) != length(y))
  X <- list(...)
  names <- names(X[[1]])
  names(names) <- names
  lapply(names, function(n) {
    unlist(lapply(X, function(x) x[[n]]))
  })
}
```

@@

The wave patterns described by these complex EOFs match the patterns seen in the *navie* EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3. Note that in Figure \@ref(fig:ceofs-2), the real part of EOF1 has non zero mean. This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean. The real part of the first EOF is, therefore, capturing the mean zonally anomalous field *and the variability that projects onto it*.

(ref:extended-series-cap) Temporal series extended using ERA5 Back extended preliminary edition (period 1950 -- 1978), ERA5 (period 1979 -- 2019) and ERA20C (period 1900 -- 2010). Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern. In numbers, the correlation between the ERA20C-based series and the two ERA5-based series. The ERA5 + ERABC and the ERA20C series are centred and scaled using the mean and standard deviation for their period of overlap (1950 -- 2010).

@@

extended <- rbindlist(list(era20c = era20c,
                           era5 = era_more),  use.names = TRUE, idcol = "dataset") %>%
  .[, hgt := (hgt - mean(hgt[dataset == "era5"], na.rm = TRUE))/sd(hgt[dataset == "era5"], na.rm = TRUE),
    by = .(part, PC)]

correlations <- extended %>% 
  dcast(time + PC + part ~ dataset, value.var = "hgt") %>% 
  na.omit() %>% 
  .[, cor(era20c, era5), 
    by = .(PC, part, time = ifelse(year(time) >= 1979, 2005, 1960))] %>% 
  .[, time := lubridate::as_datetime(paste0(time,  "-01-01"))] %>% 
  .[, dataset := ifelse(year(time) >= 1979, "era5", "era5be")]

extended %>% 
  copy() %>% 
  .[dataset == "era5", dataset := ifelse(year(time) < 1979, "era5be", "era5")] %>% 
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = dataset)) +
  geom_text(data = correlations, aes(color = dataset, y = 2.3, label = signif(V1, 2))) +
  scale_x_datetime(NULL) +
  scale_y_continuous("Standarised EOF") +
  scale_color_brewer("Data", palette = "Dark2", labels = c(era5 = "ERA5",
                                                           era20c = "ERA20C",
                                                           era5be = "ERA5 BE"),
                     guide = guide_legend(override.aes = list(size = 1), order = 1)) +
  ggnewscale::new_scale_colour() +
  geom_smooth(data = ~.x[dataset == "era20c"][, dataset := dataset == "era20c"], 
              method = "lm", se = FALSE,  aes(color = dataset)) +
  geom_smooth(data = ~copy(.x)[, dataset := dataset == "era20c"][time %between% common_range(time, dataset)],
              aes(color = dataset), method = "lm", se = FALSE) +

  scale_color_brewer("Trend", palette = "Dark2", direction = -1,
                     guide = guide_legend(override.aes = list(size = 1), order = 99),
                     labels = c("FALSE" = "ERA5+ERA5 BE",
                                "TRUE" = "ERA2CO")) +
  facet_grid(part ~ PC) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  tag_facets("cr")
```

Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 and using ERA20C for the period 1900 -- 2010. The hybrid ERA5 + ERA5 BE series and the ERE20C series agree relatively well during the period of overlap, although correlations are higher in the satellite period. The ERA20C time-series has a significant upward linear trend in the imaginary part of EOF1. However, there is no trend in the corresponding hybrid modern dataset during the period of overlap and the correlations between the old and modern series is much poorer. This suggests that the observed trends is not robust.

@@

    scale_x_longitude() +
    scale_y_latitude() +
    facet +
    coord_polar() +
    axis_labs_smol 
}

@@

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
plot_regr(regr, "PC1", c(200, 50), "hgt")
```

@@

```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)"}

time <- ceof[, eof[[1]]$left] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(PC, part)] %>%
  .[]

"/home/elio/Documents/CONICET/asymsam/analysis/data/derived_data/indexes.Rds" %>%
  readRDS() %>%
  # .[lev %in% c(200, 50)] %>%
  # .[term != "full"] %>%
  rm_singleton() %>%
  .[season(time) == "SON"] %>%
  .[, .(estimate_norm = mean(estimate_norm)), by = .(lev, term, time = seasonally(time))] %>%
  .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
  na.omit() %>%
  .[, cor(hgt, estimate_norm)^2, by = .(lev, term, part, PC)] %>%
  ggplot(aes(lev, V1)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²") +
  scale_size_manual(guide = "none",
                    values = c(full = 1,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "Asymmetric SAM",
                                                         sym = "Symmetric SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip()

>>>>

lab_lev <- AddSuffix(" hPa")
lab_cplx <- c(R = "Real", 
              I = "Imaginary")
lab_PC <- c(PC1 = "EOF1", 
            PC2 = "EOF2")

combine_lists <- function(...) {
  # browser(expr = length(x) != length(y))
  X <- list(...)
  names <- names(X[[1]])
  names(names) <- names
  lapply(names, function(n) {
    unlist(lapply(X, function(x) x[[n]]))
  })
}
```

@@

The wave patterns described by these complex EOFs match the patterns seen in the *navie* EOFs of Figure \@ref(fig: cor-eof-naive-1): The first is a wave 1, while the second is a wave 3. Note that in Figure \@ref(fig:ceofs-2), the real part of EOF1 has non zero mean. This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean (shown in Figure \@ref(fig:mean-anom)), not the time mean. The real part of the first EOF is, therefore, capturing the mean zonally anomalous field *and the variability that projects onto it*.

```{r mean-anom, fig.cap = "Mean SON zonal geopotential height anomalies at (a) 50 hPa and (b) 200 hPa"}
era5 %>% 
  .[, mean(-hgt), by = .(lon, lat, lev)] %>% 
  .[lev %in% c(50, 200)] %>% 
  .[lat <= -20] %>% 
  .[, V2 := V1 - mean(V1), by = .(lat, lev)] %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
  geom_contour_tanaka(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
  geom_qmap(~ .x[lat <= -20]) +
  scale_fill_divergent(NULL, breaks = AnchorBreaks(exclude = 0), 
                       guide = guide_colorstrip_bottom()) +
  scale_x_longitude() +
  scale_y_latitude() +
  facet_grid(~lev, labeller = labeller(lev = lab_lev)) +
  coord_polar() +
  axis_labs_smol +
  tag_facets()
```

(ref:extended-series-cap) Temporal series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019) and ERA20C (period 1900 -- 2010). Each series is computed by projecting monthly geopotential height zonal anomalies standardised by level south of 20ºS onto the corresponding spatial pattern. In numbers, the coefficient of determination between the ERA20C-based series and the two ERA5-based series. The ERA5 + ERABC and the ERA20C series are centred and scaled using the mean and standard deviation for their period of overlap (1950 -- 2010).

@@

extended <- rbindlist(list(era20c = era20c,
                           era5 = era_more),  use.names = TRUE, idcol = "dataset")

trends <- extended %>% 
  .[time %between% common_range(time, dataset)] %>% 
  .[, FitLm(hgt, time, se = TRUE), by = .(dataset, part, PC)] %>% 
  rm_intercept() %>% 
  .[, p_val := Pvaluate(estimate, std.error, df)] %>% 
  .[, p_report := report_p(p_val)]

trends_all <- extended %>% 
  .[, FitLm(hgt, time, se = TRUE), by = .(dataset, part, PC)] %>% 
  rm_intercept() %>% 
  .[, p_val := Pvaluate(estimate, std.error, df)] %>% 
  .[, p_report := report_p(p_val)]

periods <- range(extended[time %between% common_range(time, dataset)]$time)
periods <- c(periods[1], lubridate::as_datetime("1979-01-01"), periods[2])

correlations <- extended %>% 
  dcast(time + PC + part ~ dataset, value.var = "hgt") %>% 
  na.omit() %>% {
    d <- .
    rbind(
  d[, .(correlation = cor(era20c, era5)), 
    by = .(PC, part, time = ifelse(year(time) >= 1979, 
                                   as.character(mean(periods[2:3])),
                                   as.character(mean(periods[1:2]))))],

    d[, .(correlation = cor(era20c, era5)), 
    by = .(PC, part)] %>% 
    .[, time := as.character(mean(periods[c(1, 3)]))]
    )
  } %>%

  .[, time := lubridate::as_datetime(time)]

extended %>% 
  copy() %>% 
  .[, ref := time %between% common_range(time, dataset)] %>%
  .[, hgt := (hgt - mean(hgt[ref]))/sd(hgt[ref]), by = .(PC, part, dataset)] %>% 
  .[, dataset_o := dataset] %>% 
  # .[dataset == "era5", dataset := ifelse(year(time) < 1979, "era5be", "era5")] %>% 
  ggplot(aes(time, hgt)) +
  # Zero line + rangeframe
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +

  geom_line(aes(color = dataset)) +
  scale_alpha_manual(values = c(1, 0.2)) +
  # Correlations
  annotate("errorbarh", xmin = periods[1:2], xmax = periods[2:3], y = -2.7, size = 0.3) +
  annotate("errorbarh", xmin = periods[1], xmax = periods[3], y = -4, size = 0.3) +

  geom_label(data = correlations[time %in% range(time)], aes(y = -2.7, label = signif(correlation^2, 2)), 
             label.r = grid::unit(0, "lines"), label.size = 0) +
  geom_label(data = correlations[!(time %in% range(time))], aes(y = -4, label = signif(correlation^2, 2)), 
             label.r = grid::unit(0, "lines"), label.size = 0) +

  geom_smooth(data = ~.x[dataset == "era20c"], 
              method = "lm", se = FALSE,  aes(color = dataset)) +
  geom_smooth(data = ~copy(.x)[time %between% common_range(time, dataset)],
              aes(color = dataset), method = "lm", se = FALSE) +

  scale_x_datetime(NULL) +
  scale_y_continuous("Standarised EOF") +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(era20c = "ERA20C",
                                                           era5 = "ERA5 BE"),
                     guide = guide_legend(override.aes = list(size = 1), order = 1)) +
  facet_grid(part ~ PC, labeller = labeller(PC = lab_PC)) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +

  tag_facets("cr")
```

Figure \@ref(fig:extended-series) shows temporal series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950 (which we call the "hybrid ERA5" reanalysis) and using ERA20C for the period 1900 -- 2010. The hybrid ERA5 series and the ERE20C series agree relatively well during the period of overlap, although correlations are higher in the satellite period. There is a slight downwards trend in the real part of EOF1 in the hybrid reanalysis (Figure \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1" & dataset != "era20c"]$p_report` for the period 1950 -- 2010) that appears to be part of a long-term signal, judging by the statistically significant trend observed in the ERA20C reanalysis from 1900 to 2010 (p-value `r trends_all[part == "Real" & PC == "PC1" & dataset == "era20c"]$p_report`) with similar magnitude during the period of overlap.

The ERA20C time-series has a also significant upward linear trend in the imaginary part of EOF1 (Figure \@(ref:fig-extended-series)a.2, p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset == "era20c"]$p_report`. However, there is no trend in the hybrid dataset during the period of overlap (p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset != "era20c"]$p_report`) and the correlations between the old and modern series are much poorer for this variable. This suggests that the observed trend is not robust.

There is no significant trend in any of the complex parts of EOF2. The correlations between the hybrid and ERA20C datasets are much higher than for the EOF1, which suggests that ERA20C could be used as a valid dataset to study this index in the past.

<!-- (ref:extended-magnitude-cap) Same as Figure \@ref(fig:extended-series) but for EOF magnitude defined as $\sqrt{\mathrm{Real}^2 + \mathrm{Imaginary}^2}$.  -->

<!-- ```{r extended-magnitude, fig.cap = "(ref:extended-magnitude-cap)"} -->
<!-- magnitudes <- extended %>%  -->
<!--   copy() %>%  -->
<!--   dcast(time + PC + dataset ~ part, value.var = "hgt") %>%  -->
<!--   .[, ref := time %between% common_range(time, dataset)] %>% -->
<!--   .[, hgt := sqrt(Real^2 + Imaginary^2)] %>%  -->
<!--   .[, hgt := (hgt - mean(hgt[ref]))/sd(hgt[ref]), by = .(PC, dataset)] -->

<!-- correlations_mag <- magnitudes %>%  -->
<!--   dcast(time + PC  ~ dataset, value.var = "hgt") %>%  -->
<!--   na.omit() %>% { -->
<!--     d <- . -->
<!--     rbind( -->
<!--   d[, .(correlation = cor(era20c, era5)),  -->
<!--     by = .(PC, time = ifelse(year(time) >= 1979,  -->
<!--                                    as.character(mean(periods[2:3])), -->
<!--                                    as.character(mean(periods[1:2]))))], -->

<!--     d[, .(correlation = cor(era20c, era5)),  -->
<!--     by = .(PC)] %>%  -->
<!--     .[, time := as.character(mean(periods[c(1, 3)]))] -->
<!--     ) -->
<!--   } %>%  -->

<!--   .[, time := lubridate::as_datetime(time)] -->

<!-- trends_mag <- magnitudes %>%  -->
<!--   .[time %between% common_range(time, dataset)] %>%  -->
<!--   .[, FitLm(hgt, time, se = TRUE), by = .(dataset, PC)] %>%  -->
<!--   rm_intercept() %>%  -->
<!--   .[, p_val := Pvaluate(estimate, std.error, df)] %>%  -->
<!--   .[, p_report := report_p(p_val)] -->

<!-- trends_all_mag <- magnitudes %>%  -->
<!--   .[, FitLm(hgt, time, se = TRUE), by = .(dataset, PC)] %>%  -->
<!--   rm_intercept() %>%  -->
<!--   .[, p_val := Pvaluate(estimate, std.error, df)] %>%  -->
<!--   .[, p_report := report_p(p_val)] -->

<!-- magnitudes %>%  -->
<!--   ggplot(aes(time, hgt)) + -->
<!--   # Zero line + rangeframe -->
<!--   geom_hline(yintercept = 0, size = 0.1, color = "gray30") + -->
<!--   ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") + -->

<!--   geom_line(aes(color = dataset)) + -->

<!--   # Correlations -->
<!--   annotate("errorbarh", xmin = periods[1:2], xmax = periods[2:3], y = -2.7, size = 0.3) + -->
<!--   annotate("errorbarh", xmin = periods[1], xmax = periods[3], y = -4, size = 0.3) + -->

<!--   geom_label(data = correlations_mag[time %in% range(time)], aes(y = -2.7, label = signif(correlation^2, 2)),  -->
<!--              label.r = grid::unit(0, "lines"), label.size = 0) + -->
<!--   geom_label(data = correlations_mag[!(time %in% range(time))], aes(y = -4, label = signif(correlation^2, 2)),  -->
<!--              label.r = grid::unit(0, "lines"), label.size = 0) + -->

<!--     geom_smooth(data = ~.x[dataset == "era20c"],  -->
<!--               method = "lm", se = FALSE,  aes(color = dataset)) + -->
<!--   geom_smooth(data = ~copy(.x)[time %between% common_range(time, dataset)], -->
<!--               aes(color = dataset), method = "lm", se = FALSE) + -->

<!--   scale_x_datetime(NULL) + -->
<!--   scale_y_continuous("Standarised EOF magnitude") + -->
<!--   scale_color_brewer(NULL, palette = "Dark2", labels = c(era20c = "ERA20C", -->
<!--                                                            era5 = "ERA5 BE"), -->
<!--                      guide = guide_legend(override.aes = list(size = 1), order = 1)) + -->
<!--   facet_wrap(~PC, labeller = labeller(PC = lab_PC), ncol = 2) + -->
<!--   theme(panel.spacing = grid::unit(1.5, "lines"))  + -->

<!--   tag_facets() -->
<!-- ``` -->

@@

    scale_x_longitude() +
    scale_y_latitude() +
    facet +
    coord_polar() +
    axis_labs_smol +
    tag_facets("rc")
}

@@

```{r eof1-regr-gh, fig.cap = "(ref:eof1-regr-gh-cap)"}
plot_regr(regr, "PC1", c(200, 50), "hgt") 
```
Figure \@ref(fig:ref:eof1-regr-gh) shows regression patterns of SON geopotential height on the real and imaginary part of the first EOF at 50 hPa and 200 hPa.

@@

```{r sam-eof-vertical, fig.cap = "(ref:sam-eof-vertical-cap)"}
time <- ceof[, eof[[1]]$left] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(PC, part)] %>%
  .[]

sams <- readRDS(data_path("derived", "sam.Rds"))

sams %>% 
  .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
  na.omit() %>%
  .[, cor(hgt, estimate)^2, by = .(lev, term, part, PC)] %>%
  ggplot(aes(lev, V1)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²") +
  scale_size_manual(guide = "none",
                    values = c(full = 1,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "Asymmetric SAM",
                                                         sym = "Symmetric SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip()

```{r}
sams <- ERA5() %>% 
  ReadNetCDF(vars = c("hgt" = "z"), 
             subset = list(latitude = c(-90, -20))) %>% 
  normalise_coords() %>% 
  .[season(time) == "SON"] %>% 
  .[, .(hgt = mean(hgt)), by = .(lon, lat, lev, time = seasonally(time))] %>% 
  .[, hgt := Anomaly(hgt), by = .(lon, lat, lev)] %>% 
  .[, asymsam::eof_asym(hgt, lon, lat, time), by = lev]
```

```{r}
sams %>% 
  .[, PC := NULL] %>% 
 .[time, on =.NATURAL, allow.cartesian = TRUE] %>%
  na.omit() %>%
  .[, cor(hgt, estimate)^2, by = .(lev, term, part, PC)] %>%
  ggplot(aes(lev, V1)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²") +
  scale_size_manual(guide = "none",
                    values = c(full = 1,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "Asymmetric SAM",
                                                         sym = "Symmetric SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip()
